Active Learning for SAT Solver Benchmarking

### 2022-10-11 paper/paper.tex
Dissolve Section 5.1 (Runtime Prediction)
Move content to Section 4 (Experimental Design) and cut. Also moved the
runtime-discretization results from Section 3.1. Paper is roughly a page
shorter now (partly also because we integrated Algo 2 into the text
instead of having it on a separate page).

We transform the runtimes into discrete runtime labels on a per-instance basis.
For each instance $e \in \mathcal{I}$, we use a single-link hierarchical clustering algorithm to assign the runtimes in $\bigl\{ r(a, e) \mid a \in A \bigr\}$ to one of the $k$ clusters $C_1, \dots, C_k$ such that the fastest runtimes for instance $e$ are in cluster $C_1$ and the slowest are in cluster $C_k$.
For instances with many timeouts, we have to manually fix the clustering results such that all timeouts $\tau$ are in cluster $C_k$.
%TF: maybe a bit unclear. We do not manually fix anything. There is always one cluster with only time-outs if there are any.
%JB: was wondering about that as well; how do we exactly create the timeout cluster? do we only cluster with k-1, removing timeout instances, or is the timeout cluster an ordinary cluster (which might also contain on-timeout instances)?
The runtime transformation function $\gamma_k : {\mathcal{A} \times \mathcal{I}} \rightarrow \left\lbrace 1, \dots, k \right\rbrace$ is then specified as follows:
%
$$\gamma_k(a, e) = j ~\Leftrightarrow~ r(a, e) \in C_j$$

@@

However, we need to ensure that discretized labels still discriminate solvers and align with the actual PAR-2 ranking.
We analyzed the ranking induced by $s_3$ in preliminary experiments with the SAT Competition~2022 Anniversary Track~\cite{sat2022}.
According to a Wilcoxon-signed-rank test with $\alpha = 0.05$, \SI{87.83}{\%} of solver pairs have significantly different scores after discretization, only a slight drop compared to \SI{89.95}{\%} before discretization.
%While only time-out labels (solver times out or not on a particular instance) are sufficient to significantly distinguish about \SI{87.04}{\%} of solver pairs.
%\footnote{Based on hierarchical clustering with a log-single-link criterion for $k = 3$.}
Further, our ranking approach is capable of correctly deciding for almost all (about \SI{97.45}{\%}; $\sigma = \SI{3.68}{\%}$) solver pairs which solver is faster.
%Moreover, it decides all solvers correctly if the difference in the cluster-label-ranking scores between two solvers is at least \SI{0.11}{}.
%Refer to Table~\ref{tab:discreteranking} for a comparison with the PAR-2 ranking.
In particular, the Spearman correlation of $s_3$ and PAR-2 ranking is about \SI{0.988}{}, which is very close to the optimal value of 1~\cite{de2016comparing}.
All these results show that discretized runtimes are suitable for our framework.

@@

\begin{algorithm}[htbp]
  \caption{Evaluation Framework}
  \label{alg:eval}

@@

For hyper-parameter tuning, we randomly sample \SI{10}{\%} of the complete set of 5355 instances with stratification regarding the instance's family.
All instance families that are too \emph{small}, i.e., \SI{10}{\%} of them corresponds to less than one instance, are put into one meta-family for stratification.
This smaller dataset allows for a more extensive exploration of hyper-parameter space.

@@

Given Algorithm~\ref{algALBenchmark}, there are several possible instantiations for the three phases, i.e., \emph{selection}, \emph{stopping}, and \emph{ranking}.
Also, there are different choices for the runtime-label prediction model.
Note that we are not considering all previously listed approaches since a grid search of all combinations would be infeasible.
Rather, we filter approaches based on preliminary experimental results (cf. Section~\ref{sec:eval:prediction}) and do the main end-to-end experiments only with a subset.
The end-to-end experiment configurations are given below.

@@

\subsubsection{Runtime-label prediction model.}

As runtime-label prediction model, we only make use of one fixed model since an exhaustive grid search would be infeasible (cf.~Section~\ref{sec:eval:prediction}).
Our model of choice is an ensemble, stacking a quadratic-discriminant analysis\footnote{With a regularization parameter of zero and a singular value threshold of zero} onto a random forest model\footnote{With \emph{entropy} splitting criterion and \emph{balanced} class weights} using a decision tree\footnote{With \emph{Gini} impurity splitting criterion; choosing the best split for each branching decision and maximum tree depth of 5 levels} to weight its members.
If not stated explicitly, all other model parameters default to the presets of \emph{scikit-learn}\footnote{\url{https://scikit-learn.org/stable/index.html}; Version 1.0.2}~\cite{scikit-learn}.

%the proposed runtime label ranking, (2) the PAR-2 subsample ranking, and (3) the runtime label ranking that uses the PAR-2 subsample ranking as a fallback when performance differences drop below a threshold.
%Possible values are \SI{0.01}{}, \SI{0.05}{}, and \SI{0.1}{}.
%Also, we experiment with different lengths of the prediction history that is used for stabilizing the ranking.
%Possible values are 10, 20, 30, 40, and 50.

@@

Before evaluating our approach end-to-end as described in Section~\ref{sec:exdesign}, we discuss the choice of the machine-learning predictor and the runtime-discretization method.
Thereafter, we look at the results of our active-learning approach.
Finally, we also analyze the importance of different instance families to our framework.

\subsection{Runtime Prediction}
\label{sec:eval:prediction}

%JB: we should probably remove this section here and jump right into active learning;
% - results of classifier comparison can shortly be described in Section 4 (new subsection) or 3.2 ("Solver Model", which currently mainly describes runtime discretization); in any case, classifier should be discussed earlier
% - in current version, we already state in Section 3.2 that we use hierarchical clustering, so description of algorithm and some "performed well in preliminary experiments" could go there
An exhaustive grid search of all active-learning approaches outlined in Section~\ref{sec:main} and different runtime-prediction models is infeasible.
Therefore, we compare predictors first and select one runtime-prediction model for all further experiments.
We evaluate predictors with the \emph{Matthews Correlation Coefficient} (MCC) scores~\cite{gorodkin2004comparing,matthews1975comparison} of the classification task.
MCC scores are great in dealing with class imbalances in contrast to conventional metrics like accuracy.

% \begin{table}
%   \centering
%   \caption{
%   	Time-out prediction Matthews Correlation Coefficient performances.
%   	Results are obtained using cross-validation.
%   	We provide \SI{64}{\%} of ground-truth for training, use \SI{16}{\%} as a validation set, and \SI{20}{\%} as a test set (5-fold cross-validation).
%   }
%   \label{tab:mcc}
%   \begin{tabular}{lc}
%     Prediction Model & Avg. MCC (Std.) \\
%     \hline
%     Stacking ensemble (QDA, RF)           & 0.9527 (0.0292) \\
%     Quadratic-discriminant analysis (QDA) & 0.9290 (0.0339) \\
%     Random forest (RF)                    & 0.8530 (0.0479) \\
%     AdaBoost with decision trees          & 0.8384 (0.0444) \\
%     AdaBoost with logistic regression     & 0.8158 (0.0985) \\
%     Decision tree                         & 0.8059 (0.0707) \\
%     Logistic regression                   & 0.8052 (0.1018) \\
%     kNN classifier                        & 0.7885 (0.1521) \\
%     MLP classifier                        & 0.7760 (0.1408) \\
%     Support-vector classifier             & 0.7757 (0.2149) \\
%     Naive Bayes                           & 0.7306 (0.1394) \\
%     \hline
%   \end{tabular}
% \end{table}

% Stacking MCC
% 0.888221577313471
% 0.04753979056930393

%To obtain the results in Table~\ref{tab:mcc},
We perform cross-validation among all solvers: we assume that all other solver runtimes are known and provide \SI{64}{\%} of ground-truth runtime labels of the target solver for training, use \SI{16}{\%} as a validation set for hyper-parameter tuning of the runtime predictor, and \SI{20}{\%} as a test set (5-fold cross-validation).
This is repeated for each solver once.
Note that we perform cross-validation on two levels, i.e., among the solvers and the target solver runtimes.
We compare the average performances on the test sets.

ML classification models that we have tried include random forests, decision trees, quadratic-discriminant analysis, AdaBoost, logistic regression, $k$-nearest neighbor, multi-layer perceptrons, support vector machines, and naive Bayes.
A stacking ensemble consisting of quadratic-discriminant analysis and random forests performed best.
It produces an MCC score of \SI{0.8882}{} for predicting three-class runtime labels.
% \SI{0.9527}{} for time-out prediction
The results are significantly better than any other model that we have tried regarding a Wilcoxon signed-rank test with $\alpha = \SI{5}{\%}$.

%Since the differences in the MCC-performance distributions of the best and second-best approaches are statistically significant regarding a Wilcoxon signed-rank test with $\alpha = \SI{5}{\%}$, we only use the best approach for all further experiments as a prediction model.

For runtime discretization, we use hierarchical clustering with $k = 3$ and a log-single-link criterion for all further experiments.
Each non-time-out runtime starts in a separate interval.
We then gradually merge intervals whose single-link logarithmic distance is the smallest until the desired number of partitions is reached.
As already mentioned in Section~\ref{sec:main:model}, this approach has good properties regarding the predictiveness of the PAR-2 score ranking and its discriminatory power.
Also, experiments showed that our stacking ensemble performed particularly well for this kind of discretization method.
Other clustering approaches that we have tried include hierarchical clustering with mean-, median- and complete-link criteria, as well as $k$-means and spectral clustering.
%Predicting runtime labels in a similar experiment as above produced an MCC score of on average \SI{88.82}{\%} ($\sigma = \SI{4.75}{\%}$; in comparison to \SI{95.27}{\%} for predicting time-outs).
%JB: "similar experiment as above" - reference unclear; what do you want to say?
%TF: Removed the sentence

\subsection{Active-Learning Results}

\subsubsection{Tuning our algorithm.}

@@

\subsubsection{Full dataset evaluation.}

>>>>

We transform the runtimes into discrete runtime labels on a per-instance basis.
For each instance $e \in \mathcal{I}$, we use clustering algorithm to assign the runtimes in $\bigl\{ r(a, e) \mid a \in A \bigr\}$ to one of the $k$ clusters $C_1, \dots, C_k$ such that the fastest runtimes for instance $e$ are in cluster $C_1$ and the slowest are in cluster $C_k$.
For instances with many timeouts, we have to manually fix the clustering results such that all timeouts $\tau$ are in cluster $C_k$.
%TF: maybe a bit unclear. We do not manually fix anything. There is always one cluster with only time-outs if there are any.
%JB: was wondering about that as well; how do we exactly create the timeout cluster? do we only cluster with k-1, removing timeout instances, or is the timeout cluster an ordinary cluster (which might also contain on-timeout instances)?
The runtime transformation function $\gamma_k : {\mathcal{A} \times \mathcal{I}} \rightarrow \left\lbrace 1, \dots, k \right\rbrace$ is then specified as follows:
%
$$\gamma_k(a, e) = j ~\Leftrightarrow~ r(a, e) \in C_j$$

@@

@@

\begin{algorithm}[htb]
  \caption{Evaluation Framework}
  \label{alg:eval}

@@

For hyper-parameter tuning, we randomly sample \SI{10}{\%} of the complete set of 5355 instances with stratification regarding the instance's family.
All instance families that are too \emph{small}, i.e., \SI{10}{\%} of them corresponds to less than one instance, are put into one meta-family for stratification.
This \emph{tuning dataset} allows for a more extensive exploration of hyper-parameter space.

@@

Given Algorithm~\ref{algALBenchmark}, there are several possible instantiations for the three phases, i.e., \emph{selection}, \emph{stopping}, and \emph{ranking}.
Also, there are different choices for the runtime-label prediction model and runtime discretization.
Note that we are not considering all previously listed approaches since a grid search of all combinations would be infeasible.
Rather, we filter approaches based on preliminary experimental results and do the end-to-end experiments only with a subset.
The end-to-end experiment configurations are given below.

@@

\subsubsection{Prediction model.}

For runtime-label prediction, we only use one model configuration in our experiments since an exhaustive grid search would be infeasible.
In preliminary experiments, we compared a variety of model types from \emph{scikit-learn}~\cite{scikit-learn}.
In particular, we conducted nested cross-validation, including hyper-parameter tuning, and used Matthews Correlation Coefficient~\cite{gorodkin2004comparing,matthews1975comparison} to assess the performance for predicting runtime labels.
Our final choice is a stacking ensemble of a quadratic-discriminant analysis and a random forest, using a decision tree to combine its ensemble members.

%Our model of choice is an ensemble, stacking a quadratic-discriminant analysis\footnote{With a regularization parameter of zero and a singular value threshold of zero} onto a random forest model\footnote{With \emph{entropy} splitting criterion and \emph{balanced} class weights} using a decision tree\footnote{With \emph{Gini} impurity splitting criterion; choosing the best split for each branching decision and maximum tree depth of 5 levels} to weight its members.
%If not stated explicitly, all other model parameters default to the presets of \emph{scikit-learn}\footnote{\url{https://scikit-learn.org/stable/index.html}; Version 1.0.2}~\cite{scikit-learn}.
%
%
%Therefore, we compare predictors first and select one runtime-prediction model for all further experiments.
%We evaluate predictors with the \emph{Matthews Correlation Coefficient} (MCC) scores~\cite{gorodkin2004comparing,matthews1975comparison} of the classification task.
%MCC scores are great in dealing with class imbalances in contrast to conventional metrics like accuracy.
%
%% \begin{table}
%	%   \centering
%	%   \caption{
%		%   	Time-out prediction Matthews Correlation Coefficient performances.
%		%   	Results are obtained using cross-validation.
%		%   	We provide \SI{64}{\%} of ground-truth for training, use \SI{16}{\%} as a validation set, and \SI{20}{\%} as a test set (5-fold cross-validation).
%		%   }
%	%   \label{tab:mcc}
%	%   \begin{tabular}{lc}
%		%     Prediction Model & Avg. MCC (Std.) \\
%		%     \hline
%		%     Stacking ensemble (QDA, RF)           & 0.9527 (0.0292) \\
%		%     Quadratic-discriminant analysis (QDA) & 0.9290 (0.0339) \\
%		%     Random forest (RF)                    & 0.8530 (0.0479) \\
%		%     AdaBoost with decision trees          & 0.8384 (0.0444) \\
%		%     AdaBoost with logistic regression     & 0.8158 (0.0985) \\
%		%     Decision tree                         & 0.8059 (0.0707) \\
%		%     Logistic regression                   & 0.8052 (0.1018) \\
%		%     kNN classifier                        & 0.7885 (0.1521) \\
%		%     MLP classifier                        & 0.7760 (0.1408) \\
%		%     Support-vector classifier             & 0.7757 (0.2149) \\
%		%     Naive Bayes                           & 0.7306 (0.1394) \\
%		%     \hline
%		%   \end{tabular}
%	% \end{table}
%
%% Stacking MCC
%% 0.888221577313471
%% 0.04753979056930393
%
%%To obtain the results in Table~\ref{tab:mcc},
%We perform cross-validation among all solvers: we assume that all other solver runtimes are known and provide \SI{64}{\%} of ground-truth runtime labels of the target solver for training, use \SI{16}{\%} as a validation set for hyper-parameter tuning of the runtime predictor, and \SI{20}{\%} as a test set (5-fold cross-validation).
%This is repeated for each solver once.
%Note that we perform cross-validation on two levels, i.e., among the solvers and the target solver runtimes.
%We compare the average performances on the test sets.
%
%ML classification models that we have tried include random forests, decision trees, quadratic-discriminant analysis, AdaBoost, logistic regression, $k$-nearest neighbor, multi-layer perceptrons, support vector machines, and naive Bayes.
%A stacking ensemble consisting of quadratic-discriminant analysis and random forests performed best.
%It produces an MCC score of \SI{0.8882}{} for predicting three-class runtime labels.
%% \SI{0.9527}{} for time-out prediction
%The results are significantly better than any other model that we have tried regarding a Wilcoxon signed-rank test with $\alpha = \SI{5}{\%}$.
%%Since the differences in the MCC-performance distributions of the best and second-best approaches are statistically significant regarding a Wilcoxon signed-rank test with $\alpha = \SI{5}{\%}$, we only use the best approach for all further experiments as a prediction model.

\subsubsection{Runtime discretization.}

To define prediction targets, i.e., discretized runtime labels, we use hierarchical clustering with $k = 3$ and a log-single-link criterion, which produced the most \emph{useful} labels in preliminary experiments.
We denote this adapted solver scoring function with $s_3$.
Other clustering approaches that we have tried include hierarchical clustering with mean-, median- and complete-link criteria, as well as $k$-means and spectral clustering.
%Predicting runtime labels in a similar experiment as above produced an MCC score of on average \SI{88.82}{\%} ($\sigma = \SI{4.75}{\%}$; in comparison to \SI{95.27}{\%} for predicting time-outs).
%JB: "similar experiment as above" - reference unclear; what do you want to say?
%TF: Removed the sentence
In our chosen hierarchical procedure, each non-time-out runtime starts in a separate interval.
We then gradually merge intervals whose single-link logarithmic distance is the smallest until the desired number of partitions is reached.

To obtain \emph{useful} labels, we need to ensure that discretized labels still discriminate solvers and align with the actual PAR-2 ranking.
We analyzed the ranking induced by $s_3$ in preliminary experiments with the SAT Competition~2022 Anniversary Track~\cite{sat2022}.
According to a Wilcoxon-signed-rank test with $\alpha = 0.05$, \SI{87.83}{\%} of solver pairs have significantly different scores after discretization, only a slight drop compared to \SI{89.95}{\%} before discretization.
%While only time-out labels (solver times out or not on a particular instance) are sufficient to significantly distinguish about \SI{87.04}{\%} of solver pairs.
%\footnote{Based on hierarchical clustering with a log-single-link criterion for $k = 3$.}
Further, our ranking approach correctly decides for almost all (about \SI{97.45}{\%}; $\sigma = \SI{3.68}{\%}$) solver pairs which solver is faster.
%Moreover, it decides all solvers correctly if the difference in the cluster-label-ranking scores between two solvers is at least \SI{0.11}{}.
%Refer to Table~\ref{tab:discreteranking} for a comparison with the PAR-2 ranking.
In particular, the Spearman correlation of $s_3$ and PAR-2 ranking is about \SI{0.988}{}, which is very close to the optimal value of 1~\cite{de2016comparing}.
All these results show that discretized runtimes are suitable for our framework.

@@

In this section, we evaluate our active-learning framework.
First, we analyze and tune the different components of our framework on the tuning dataset.
Next, we evaluate the best configuration with the full dataset.
Finally, we analyze the importance of different instance families to our framework.

\subsection{Hyper-Parameter Analysis}

@@

\subsection{Full-Dataset Evaluation}
